Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists

### 2024-04-25 paper.Rmd
Final batch up edits
Please search for !!! to see comments

: Conferences, events and workshops at which the tutorials were delivered and tested.

@@

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research. This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM are introduced and the first concepts within NetLogo are explained. The third lesson aims to explain how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial build a foundation for working in NetLogo and with ABM. All activity stays within the intermediate level, although higher levels might be touched upon.

In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of homo sapiens. The learner will work with basic NetLogo syntax and learn how to set up a simulation and visualize its outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency. The learning curve is relatively gentle. This is achieved by learning to work with the NetLogo web interface. The primitives are introduced and initialization phase of a simple ABM is gone through. The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach. The learners learn about simple simulation loops and how to use primitives. The advanced level is achieved with the custom procedures and variables. In addition, if-statements are introduced and more complex versions thereof. Exporting information using plots is also explained.

In the third tutorial, the learner will build a simple trade model, picking up more advanced NetLogo coding along the way, such as loops, lists and reporters. The learner will be introduced to some techniques like modular code development and debugging which will become important with this increased coding complexity. This tutorial consists of 7 lessons. This tutorial intends to guide the learner from the intermediate to the advanced and highly specialised level of proficiency. This is achieved by improving development through using modular code, pseudocode and annotating the code. In addition, custom agent breeds are introduced. Visualization with labels and reporters, plots and monitors are learned. The use of loops is further explained. Debugging is learned in lesson 7. This can be very advanced, since it involves knowledge of the NetLogo-language combined with problem solving of unexpected behaviour.

In tutorial 4 the learner will work with Sugarscape simulations to further expand their ABM and NetLogo skills. The learner will learn how to set up more complex interactions between agents and the environment. Furthermore, the basics of setting up good experiments and validating models will be explained. This tutorial consists of 8 lessons. This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency. In this tutorial more complex agent-environment interaction is further learned, including ways to visualize the environment and how to give more agency to patches. Creating toy landscapes is central to lesson 5. The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results. The validation of agent-based models is an important concept of designing a good experiment. The finale step, is to compare simulation results with the archaeological record. The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.

In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research. This tutorial focusses less on programming in NetLogo and more on the model development process. The learner will learn about the different phases in modelling and how to export data from models. This tutorial consists of 5 lessons. This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner leans about the model development process starting with the conceptual phase of the model development. Good and bad research questions for modelling are explained and the difference between them. The learner needs to understand how to pick the right modelling technique. The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development. In this phase parametrisation, designing experiments and the analysis and interpretation of the models are important. Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication. In addition, learning how to export results from NetLogo using basic export primitives is explained. The BehaviorSpace is explained to enable the learners to analyse the output of models.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality). The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.

@@

Most of the respondents had several skills with computer applications in archaeology (see Figure \@ref(fig:computer-skills)), and many of them did have some knowledge of ABM, although a large group was completely new to the subject. The majority had never applied ABM before participating in the workshops with 14 having applied ABM. It is interesting to note that many respondents did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)). While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient. A minority rated the available theory on ABM as more than sufficient. This clearly shows the need for better educational material.

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials. Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later instalments.

@@

A large proportion of the participants of the workshops gave us information using the survey after the workshops (171 of 368 participants), a similar proportion as those responding to the survey before the workshop.

@@

The majority of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were even rated better than the teaching material, with more respondents rating them as excellent (see Figure \@ref(fig:rating-teaching)).

@@

Two more open questions were asked to the respondents. The first was aimed at learning what aspects the participants liked in the tutorials, and the other was aimed at getting feedback to improve the tutorials.

The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers very much. The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often. This was also observed during the workshops by us, most participants were working in the own pace and did not need much help. From a didactic point of view, the self-paced going through the tutorials is very efficient for teachers.

We also received feedback on possible improvements. For the first workshops we mostly received feedback on the number of bugs. During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress. The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more. Some wanted more cooperation, which is possible, but harder for some online events, although we provided break-out rooms. It is also interesting to note that over time more and more people did not see any room for improvement.

The majority of the respondents wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)). Some of the respondents were already using ABM. A large group was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM. Various participants mentioned movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence area's were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or even specialised level.

For competence area 1. Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content. The users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM. For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).

@@

Competence Area 3. Digital content creation is one of the main competence areas for the ABM-teaching material. In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content. The material does not touch upon 3.3 Copyright And Licences, but we assume that users have a general understanding. A very important aspect of learning ABM is 3.4 Programming. The users are exploring the possibilities and chances of programming an ABM. The users are expected to learn both syntax and more general concepts such as modular code development, loops, lists and commenting and documenting code.

Competence Area 5. Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving. In addition, the creation of a highly complex ABM consists of problem solving all the time, including debugging. All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves. In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and to assess their knowledge and skills.

@@

The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@jutte2024]. Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, a website has been launced ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}). On this website, the tutorials can be followed without the need to download any software.

Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources people can use to learn agent-based modelling. The website was mainly developed by one of the student groups from Saxion University of Applied Sciences, see the acknowledgements.

@@

Funding acknowledgements for the publication: The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382).

>>>>

: Events at which the tutorials were delivered and tested.

@@

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research. This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.

In tutorial 2 the user learns the basics of programming in NetLogo and develops their first simulation based on a previously published model of the Out of Africa dispersal of *Homo sapiens !!! Pls insert the following reference*

Young, David A., and Robert L. Bettinger. 1992. “The Numic Spread: A Computer Simulation.” *American Antiquity* 57 (1): 85–99. <https://doi.org/10.2307/2694836>.

. The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial intoduces several most commonly used built-in functions in Netlogo (called *primitives*) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop. It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation !!! pls insert the following reference:

Romanowska, Iza. 2018. “‪Using Agent-Based Modelling to Infer Economic Processes in the Past‬.” In *‪Cuantificar Las Economías Antiguas: Problemas y Métodos*, Publicacions i Edicions de la Universitat de Barcelona, 107–18. Barcelona: Universitat de Barcelona. <https://scholar.google.com/citations?view_op=view_citation&hl=en&user=tdcyWfoAAAAJ&citation_for_view=tdcyWfoAAAAJ:YOwf2qJgpHMC>.

Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters. The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of 7 lessons. This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with a seminal Sugarscape model !!! pls insert the following ref.

Epstein, Joshua M., and Robert L. Axtell. 1996. *Growing Artificial Societies. Social Science from the Bottom Up*. Boston: MIT Press.

to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of 8 lessons. This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.

In the final and fifth tutorial the participant learn smore about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The learner learns about the different phases of the model development process. This tutorial consists of 5 lessons. This tutorial summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results. There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often prove challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known. Finally, the learners explores the fundamental importance of replication for advancing scientific knowledge.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from !!! COMMENT PLS ADD THE NUMBER countries that represented almost all continents (see Figure \@ref(fig:nationality). There was a clear skew towards participants from !!! COMMENT what part of the world? Europe? South America. The number of women and men in our sample were close to equal (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. COMMENT - in the last sentence, are these significant differences? And if so, please provide the possible reasons for these differences.

@@

Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example !!! COMMENT pls add the most popular ones (see Figure \@ref(fig:computer-skills)), and !!! COMMENT PLS insert the % number here had some prior knowledge of ABM, while !!! COMMENT PLS insert the % number here was completely new to the subject. The majority !!! COMMENT PLS insert the % number here had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM. It is interesting to note that many respondents (!!! COMMENT PLS insert the % number here) did know what kind of software was available for ABM (see Figure \@ref(fig:abm-knowledge)).

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the availability of ABM educational resources (Figure \@ref(fig:available-theory)). While *!!! COMMENT PLS insert the % number here* of them had no opinion on the subject, a !!! COMMENT PLS insert the % number here proportion of the responded answered that they rated the available theory on ABM as limited and only a smaller group !!! COMMENT PLS insert the % number here as sufficient or as more than sufficient. This clearly shows the need for more and better educational material.

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials. Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

A similar proportion of the participants of the workshops gave us information after the workshops (171 of 368 participants).

@@

*!!! COMMENT PLS insert the % number here* of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

Two open questions were asked to the respondents: *!!! COMMENT PLS spell out the questions*

The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers very much. The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often. This was also observed during the workshops as most participants were working in the own pace and did not need much help. From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.

We also received feedback on possible improvements. During the first, big event CAA-conference in Amsterdam in April 2023 we did the first test of the tutorials and we also instructed the participants that this was work in progress. For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour. The number of references to bugs reduced to zero by the end of the project and participants started providing other feedback, for example the wish to see more examples or expanding the tutorials further. Some wanted to introduce more cooperation between participants, which we did not consider primarily aiming the tutorials for individual work. A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials may deliver solutions. It was also interesting to note that over time more and more people did not see any additional room for improvement indicating that the level, content and format of the tutorials met their needs.

We also asked the participants about their intentions in terms of ABM and their professional life. The majority of the respondents - 88%, wanted to apply ABM in the future in their research (Figure \@ref(fig:future-abm)). Some of the respondents, *!!! COMMENT PLS insert the % number here,* were already using ABM in their work. A large group *!!! COMMENT PLS insert the % number here* was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence area's were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or specialised level.

For competence area 1. Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content. In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM. For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).

@@

Competence Area 3. Digital content creation is one of the main competence areas for the ABM-teaching material. In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content. The material does not touch upon 3.3 Copyright And Licences, as they fall out of the scope of this project. A very important aspect of learning ABM is 3.4 Programming. The users are guidend through the process of developing the necessary programming skills to build a simulation. While the programming language used - NetLogo - is specific to ABM, its syntax follow general programming patterns and the learning is thus transferable to other programming languages. The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.

Competence Area 5. Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials. In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and high level understanding of the scientific process. All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models. In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial force the user to assess their knowledge and skills.

@@

The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@jutte2024]. Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) can be followed without the need to download any software.

Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources people can use to learn agent-based modelling. The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.

@@

Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union’s Horizon 2020 Research and
Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513 and the
Aarhus University Research Foundation.
